Geometric and physical interpretation of the action principle

### 2023-06-27 Articles/2022-ActionPrinciple/ScientificReportsAmended/ActionPrincipleBody.tex
ActionPrinciple: Scientific Reports version

\documentclass[10pt,twocolumn, nofootinbib]{revtex4-2}
%\documentclass[aps,pra,10pt,twocolumn,floatfix,nofootinbib]{revtex4-1}
%\documentclass[10pt,twocolumn,letterpaper]{article}

\usepackage{assumptionsofphysics}
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{
	colorlinks=true,
	citecolor=blue,
	urlcolor=blue,
	linkcolor=blue
}
\urlstyle{same}
\frenchspacing

@@

\begin{document}

\title{Geometric and physical interpretation of the action principle}
\author{Gabriele Carcassi, Christine A. Aidala}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}

@@

\section{Introduction}

While the principle of stationary action is regarded by many as one of the most important tools in physics,\cite{hamilton1834general, heaviside1903, feynman1942principle, brenier1989least, rojo2018principle} its physical meaning is not completely clear. First of all, the typical characterization of the Lagrangian as the difference between kinetic and potential energy fails even for simple systems, like a charged particle under a magnetic field. Moreover, the Lagrangian for a system is not uniquely defined, making the actual value of the action for a path not directly physically significant.  We are left to wonder: what exactly is the action and why is it stationary for actual trajectories?

@@

The mathematics needed to run the argument is well established\cite{souriau1970structure, abraham1978foundations,arnold1989mathematical, marsden1999introduction}. The two aspects that have been sorely lacking, and that we provide, are (1) a clear geometric interpretation of the action principle and (2) a tight connection between the math and the physics it represents. To make the result accessible to the widest audience, we first cover the case of a single degree of freedom using standard vector calculus. We then proceed to the case of multiple independent degrees of freedom using tools from differential geometry. The main article will presents all the key point needed to follow the argument and its physical and geometric meaning, leaving the mathematical details and calculations to the appendix.

\section{One degree of freedom}

@@

Under the assumption that
\begin{align}\label{assum_detrev}
	\tag*{(DR)}
	\parbox{2.8in}{\emph{the system undergoes deterministic and reversible evolution}}
\end{align}
we can define a displacement vector field
\begin{equation}\label{sdof_displacement}
\begin{aligned}
	\vec{S} &= \left[ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right] \\
	&= S^a e_a = \frac{d\xi^a}{dt} e_a .
\end{aligned}
\end{equation}
that describes how states move in time.\footnote{Where possible, we will be writing the same expression in both vector calculus and component notations.} The time component of the displacement vector field is constrained, as we have
\begin{equation}\label{sdof_time_constraint}
	S^t=\frac{dt}{dt}=1.
\end{equation}

If assumption \ref{assum_detrev} is valid, we expect the flow of states through a closed surface to be zero: as many states flow in as flow out of the region. Alternatively, if we assign a probability, or probability density, to each trajectory, the assumptions requires that probability not to change, so integrating the probability over a closed surface must yield zero. However we see it, assumption \ref{assum_detrev} means the field is divergence-free,\footnote{Given that this is a three dimensional space, we can use the standard tools of vector calculus.} that is
\begin{equation}\label{sdof_div_free}
	\nabla \cdot \vec{S} = \partial_a S^a = 0.
\end{equation}

@@

Because the displacement field must satisfy \ref{sdof_time_constraint}, without loss of generality we can set
\begin{equation}\label{sdof_potential_expression}
\begin{aligned}
	\vec{\theta} &= [p, 0, -H(q,p,t)] \\
	&= p e^q - H(q,p,t) e^t,
\end{aligned}
\end{equation}
where $H$ is a suitable function of $q$, $p$ and $t$. By applying definition \ref{sdof_displacement} and expanding \ref{sdof_displacement_potential} with \ref{sdof_potential_expression}, we have
\begin{equation}\label{sdof_Ham_eq}
	\left[ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right] = - \nabla \times \vec{\theta} = \left[ \frac{\partial H}{\partial p},-\frac{\partial H}{\partial q}, 1 \right],
\end{equation}
which yields Hamilton's equations. Note that the argument works in reverse: any Hamiltonian system with one degree of freedom yields a divergence-free displacement field, and therefore satisfies \ref{assum_detrev}.

@@

We now turn to constructing the principle of stationary action. As illustrated in Fig. \ref{fig_action}a, take a path $\gamma$ with endpoints $A$ and $B$, not necessarily a solution of the equations of motion. Then take a variation $\gamma'$ of that path and identify a surface $\Sigma$ between them. We can ask: what is the flow of the displacement field $\vec{S}$ through $\Sigma$? Because $\vec{S}$ is divergence-free, the flow through $\Sigma$ will depend only on the contour, therefore the question is well posed. Using Stokes' theorem, we find
\begin{equation}\label{sdof_action}
\begin{aligned}
	- \iint_{\Sigma} \vec{S} \cdot d\vec{\Sigma} &= \iint_{\Sigma} \left( \nabla \times \vec{\theta} \right) \cdot d\vec{\Sigma} \\
	=  \oint_{\partial \Sigma = \gamma \cup \gamma'} \vec{\theta}  \cdot d\vec{\gamma} 
	&= \delta \int_{\gamma} \vec{\theta} \cdot d\vec{\gamma}.
\end{aligned}
\end{equation}

@@

\section{Multiple degrees of freedom}

@@

The flow of $\vec{S}$ will still be divergence-free, but this property only tells us that the total number of states is conserved. We also need to assume that
\begin{align}\label{assum_indep}
	\tag*{(IND)}
	\parbox{2.8in}{\emph{the degrees of freedom are independent.}}
\end{align}
This means that, at each moment in time, the total number of states (i.e. the volume) of a parallelepiped will be the product of states identified by each degree of freedom (i.e. the areas). To be able to quantify the number of states identified by each degree of freedom, we introduce the rank-2 tensor (more precisely a two-form)
\begin{equation}\label{mdof_form}
	\omega = \omega_{ab} \, e^a \otimes e^b,
\end{equation}
which we call the state counting form.\footnote{Note that this approach is consistent with statistical mechanics\cite{peliti2011statistical}.}

@@

The form $\omega$ will need to be anti-symmetric
\begin{equation}\label{mdof_form_antisymm}
	\omega(\vec{v}, \vec{w}) = - \omega(\vec{w}, \vec{v})
\end{equation}
as the parallelogram identified by $\vec{v}$ and $\vec{w}$ in that order will be the same as the one identified by $\vec{w}$ and $\vec{v}$ with opposite orientation. The form $\omega$ will also be closed, meaning\footnote{In our notation, $d\Sigma$ represents the infinitesimal surface element of integration, which is the argument of the form.}
\begin{equation}\label{mdof_closed_form}
	\oiint_\Sigma \omega(d\Sigma) = 0
\end{equation}
over all closed contractible surfaces. This stems from \ref{assum_indep}. Suppose you take a surface and translate it along another independent variable. The count of states cannot depend on the independent variable. Therefore if we construct any parallelepiped at equal time, two opposite sides will contain the same number of states with opposite orientation, and therefore the integral of $\omega$ over the whole surface will be zero.

Since $\omega$ is closed, in every contractible region it can be expressed as the exterior derivative\footnote{We write the exterior derivative as $\partial \omega$ instead of $d \omega$ so that $d$ is only used for infinitesimal line or surface elements.} of a covector $\theta$:
\begin{equation}\label{mdof_form_potential}
	\begin{aligned}
		\theta &= \theta_a e^a = \theta_{q^i} e^{q^i} + \theta_{p_i} e^{p_i} + \theta_t e^t \\
		\omega &= - \partial \theta = - \left( \partial_a \theta_b - \partial_b \theta_a \right) e^a \otimes e^b
	\end{aligned}
\end{equation}
The minus sign is required to match conventions.

@@

Similarly to the single d.o.f., without loss of generality we can set
\begin{equation}\label{mdof_potential_expression}
\begin{aligned}
	\theta &= [p_i, 0, -H] \\
	&= p_i e^{q^i} - H e^t.
\end{aligned}
\end{equation}
and calculate the components of the counting form
\begin{equation}\label{mdof_form_components}
	\omega_{ab} = (-\partial\theta)_{ab} = \begin{bmatrix}
		0 & \delta^j_i & \partial_{q^i} H \\
		-\delta^i_j & 0 & \partial_{p_i} H \\
		-\partial_{q^j} H & -\partial_{p_j} H & 0
	\end{bmatrix}.
\end{equation}

@@

\section{Discussion and conclusions}

@@

Another interesting element is that a version of the principle of stationary actions holds even if \ref{assum_kineq} does not apply. In this case the ``Lagrangian'' $L(q,\dot{q},p,t)$ depends on conjugate momentum as well. Also note that the cases where the Lagrangian does not admit a Hamiltonian are exactly the cases where the action principle fails to yield a unique solution. In this sense, the action principle is better understood as a feature of Hamiltonian mechanics, instead of Lagrangian mechanics\cite{souriau1970structure, arnold1989mathematical}.

@@

@@

\section*{Appendix}

In this section we will review the arguments in more detail. Given that this article is still aimed at a physics audience, the focus will be conceptual rigor, rather than mathematical rigor. What is important is to establish a precise map between physical concepts and their mathematical representation, so that it is clear what physical requirement is captured by each mathematical definition.

We divide the appendix into three sections: the first will address the single d.o.f. case, the second will give a geometric interpretation of the tools of differential geometry that is suitable for physicists, and lastly we address the multiple d.o.f. case.

\subsection*{One degree of freedom}

For the case of one degree of freedom we are going to use vector calculus. This is possible since the problem can be studied with three variables (position, momentum and time)
\begin{equation}
\tag*{(\ref{sdof_variables})}
	\xi^a = [ q, p, t]
\end{equation}
which is the only setting where vector calculus truly works. In this sense, nature gave us a hand as three dimensional spaces are much more intuitive for us, and we can therefore develop the ideas in a familiar context before generalizing.

We break down the problem into sequential steps, which serves as a summary for the whole argument. We will have to:
\begin{enumerate}[label=(\roman*)]
	\item show that assuming determinism and reversibility \ref{assum_detrev} is equivalent to assuming the existence of the displacement field $\vec{S}$ defined in \ref{sdof_displacement} and its being divergence-free as in \ref{sdof_div_free}
	\item show that the vector potential $\vec{\theta}$ for $\vec{S}$ defined in \ref{sdof_displacement_potential} can be expressed, without loss of generality, as \ref{sdof_potential_expression}
	\item show that \ref{sdof_div_free} and \ref{sdof_potential_expression} recover Hamilton's equations as in \ref{sdof_Ham_eq}
	\item show that the flow of $\vec{S}$ through a surface $\Sigma$ delimited by a path $\gamma$ and its variation $\gamma'$ is the variation of the integral of the vector potential, as claimed in \ref{sdof_action}
	\item show that the integral of the vector potential is stationary for and only for the solutions of the equations of motion
	\item show that the integral of the vector potential reduces to the action if and only if \ref{assum_kineq} applies.
\end{enumerate}

For (i), determinism and reversibility means that giving the state at some point in time means constraining the state at all times, that the system can evolve in only one way. In math terms, if $X$ is the manifold of all possible states at all times, given $P = \{q_0, p_0, t_0\}$ there will be only one possible evolution $\gamma : \mathbb{R} \to X$ such that $\gamma(t_0) = P$.

If $X$ were a discrete space, the previous mathematical condition would be sufficient, but on a manifold it is not. Consider a damped harmonic oscillator: it satisfies the previous condition, yet the system will concentrate at the equilibrium. What needs to happen is that densities also need to be mapped exactly. That is, if a certain fraction of the system is in a given initial state, the same fraction of the system must be in the corresponding final state. Equivalently, if the system has a given probability of starting in a given initial state, it must have the same probability of ending in the corresponding final state. In mathematical terms, if $\rho : X \to \mathbb{R}$ is a statistical distribution (or a probability density) defined on our extended phase space and $\gamma$ is a possible evolution, we must have $\rho(\gamma(t_0)) = \rho(\gamma(t_1))$.

This requirement tells us two things. First, that $X$ must be a differentiable manifold. To define densities over a set of variables and map them backward and forward in time Jacobians must be defined, which means the derivative across state variables must be defined as well, in particular with respect to time, which plays the double role of a coordinate of $X$ and the evolution parameter. This means that the evolutions must be differentiable curves and therefore the displacement $\vec{S}$ must be defined, which leads to
\begin{equation}
\tag*{(\ref{sdof_displacement})}
	\begin{aligned}
		\vec{S} &= \left[ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right] \\
		&= S^a e_a = \frac{d\xi^a}{dt} e_a .
	\end{aligned}
\end{equation}
The double role of time gives us
\begin{equation}
\tag*{(\ref{sdof_time_constraint})}
 	S^t=\frac{dt}{dt}=1.
 \end{equation}

Secondly, since the densities must be constant over evolutions, the flow of these densities through a closed surface in $X$ must be zero: each evolution will enter and exit the region delimited by the surface, giving no net contribution. Mathematically this requires the displacement to be divergence-free, which leads to
\begin{equation}
\tag*{(\ref{sdof_div_free})}
	\nabla \cdot \vec{S} = \partial_a S^a = 0.
\end{equation}

For (ii), since $\vec{S}$ is divergence-free, we can find a vector potential
\begin{equation}
\tag*{(\ref{sdof_displacement_potential})}
	\begin{aligned}
		\vec{\theta} &= [\theta_q, \theta_p, \theta_t] \\
		&= \theta_a e^a \\
		\vec{S} &= - \nabla \times \vec{\theta} \\
		&= - \epsilon^{abc} \partial_b \theta_c \, e_a. \\
	\end{aligned}
\end{equation}
Note that the vector potential is not uniquely identified, since $\nabla \times(\vec{\theta} + \nabla f) = \nabla \times \vec{\theta}$ and therefore the displacement field $\vec{S}$ remains unchanged. This is commonly referred to as gauge freedom in physics, and it is analogous to what happens for a magnetic field or for an incompressible fluid.

We use this gauge freedom to set the momentum component to zero. In fact, we can always choose $f(q,p,t)$ such that $\partial_p f = -\theta_p$ and redefine $\vec{\theta}$ to be $\vec{\theta} + \nabla f$. Therefore we can set
\begin{equation}
	\vec{\theta} = [\theta_q, 0, \theta_t]
\end{equation}
without loss of generality.

We now use constraint \ref{sdof_time_constraint} on the time component. We have
\begin{align}
	S^t = \frac{dt}{dt} &= - \left(\frac{\partial}{\partial q}  \theta_p - \frac{\partial}{\partial p}  \theta_q\right) \\
	&= - \left(\frac{\partial}{\partial q}  0 - \frac{\partial}{\partial p}  \theta_q\right) \\
	& = \frac{\partial \theta_q}{\partial p} = 1
\end{align}
Integrating, we have $\theta_q = p + g(q,t)$ where $g(q,t)$ is an arbitrary function. This function can be set to zero without loss of generality as it can be removed with a gauge transformation where $f(q,t)$ does not depend on $p$, and therefore $\theta_p$ will remain unchanged. Therefore we have:
\begin{equation}
	\vec{\theta} = [p, 0, \theta_t]
\end{equation}
Lastly, we rename the last component as $\theta_t = -H$ which leads to 
\begin{equation}
\tag*{(\ref{sdof_potential_expression})}
	\begin{aligned}
		\vec{\theta} &= [p, 0, -H] \\
		&= p e^q - H e^t.
	\end{aligned}
\end{equation}

For (iii), we expand each component of \ref{sdof_displacement_potential} using \ref{sdof_potential_expression}. Again note the double role of time: $\frac{dp}{dt}$ is not the same as $\frac{\partial p}{\partial t}$.\footnote{We could be more precise and use $\tau$ for the evolution parameter, leaving $t$ just for the coordinate. The treatment would automatically acquire a relativistic flavor and would enlighten us to other interesting results, though it would distract from the main focus of this paper, the action principle. } In the first case we are taking a total derivative along the evolution, and therefore the momentum can change. In the second case we are taking a partial derivative at constant $q$ and $p$ by definition, and therefore $\frac{\partial p}{\partial t}=0$. We have
\begin{equation}
\begin{aligned}
	S^q = \frac{dq}{dt}
	&= - \left( \frac{\partial}{\partial p} \theta_t - \frac{\partial}{\partial t} \theta_p \right) \\
	&= - \left( \frac{\partial}{\partial p} (-H) - \frac{\partial}{\partial t} 0 \right) \\
	& = \frac{\partial H}{\partial p}
\end{aligned}
\end{equation}
\begin{equation}
\begin{aligned}
	S^p = \frac{dp}{dt}
	&= - \left( \frac{\partial}{\partial t} \theta_q - \frac{\partial}{\partial q} \theta_t \right) \\
	&= - \left( \frac{\partial}{\partial t} p - \frac{\partial}{\partial q} (-H) \right) \\
	& = - \frac{\partial H}{\partial q}
\end{aligned}
\end{equation}
\begin{equation}
\begin{aligned}
	S^t = \frac{dt}{dt}
	&= - \left( \frac{\partial}{\partial q} \theta_p - \frac{\partial}{\partial p} \theta_q \right) \\
	&= - \left( \frac{\partial}{\partial q} 0 - \frac{\partial}{\partial p} p \right) \\
	& = 1
\end{aligned}
\end{equation}
We have recovered Hamilton's equations
\begin{equation}
\tag*{(\ref{sdof_Ham_eq})}
\begin{aligned}
	\frac{dq}{dt} &= \frac{\partial H}{\partial p} \\
	\frac{dp}{dt} &=-\frac{\partial H}{\partial q}
\end{aligned}
\end{equation}
as the equations for a deterministic and reversible system. The only thing that the equations say is that states move in time at the same rate (i.e. $\frac{dt}{dt} = 1$) with an incompressible flow (i.e. no states are created or destroyed). That is the whole physical and geometric content of those equations.

Note that if we started from Hamilton's equations, the system would be deterministic and reversible: position and momentum at a fixed time would indeed allow us to pick a single differentiable evolution and, by Liouville's theorem, densities would be constant along such evolution. Therefore we have an equivalence between assumption \ref{assum_detrev} and Hamilton's equations \ref{sdof_Ham_eq} in the single degree of freedom case.

Now that we have developed a clear geometric and physical understanding of Hamilton's equations, we can concentrate on the action principle. For (iv) we can apply Stokes' theorem. We take a path $\gamma$ and a variation $\gamma'$. Since they share the endpoints, taken together they form a closed curve where $\gamma'$ is taken with the opposite orientation. Because $\vec{S}$ is divergence-free, the flow of $\vec{S}$ through a surface depends only on its contour. Moreover, by Stokes' theorem, the flow through the surface will be equal to the integral of the potential over the contour. Putting it all together we have
\begin{equation}
\tag*{(\ref{sdof_action})}
\begin{aligned}
	- \iint_{\Sigma} \vec{S} \cdot d\vec{\Sigma} &=
	\iint_{\Sigma} \left( \nabla \times \vec{\theta} \right) \cdot d\vec{\Sigma} \\
	&= \oint_{\partial \Sigma} \vec{\theta}  \cdot d\vec{\gamma} \\
	&= \int_{\gamma} \vec{\theta} \cdot d\vec{\gamma} - \int_{\gamma'} \vec{\theta} \cdot d\vec{\gamma}' \\
	&= \delta \int_{\gamma} \vec{\theta} \cdot d\vec{\gamma}.
\end{aligned}
\end{equation}

For (v), note that in the neighborhood of $\gamma$ the infinitesimal surface element $d\vec{\Sigma}$ has one side along $\gamma$ itself. Therefore we can write $d\vec{\Sigma} = d\vec{\gamma} \times d\vec{\lambda}$, where $d\vec{\lambda}$ is the infinitesimal displacement that connects $\gamma$ with $\gamma'$. The integrand then becomes the triple product
\begin{equation}
	\vec{S} \cdot d\vec{\gamma} \times d\vec{\lambda}.
\end{equation}
We can understand it geometrically as the volume element given by the displacement $\vec{S}$, the infinitesimal line element $d\vec{\gamma}$ and the infinitesimal variation $d\vec{\lambda}$ that moves $\gamma$ to $\gamma'$.

If $\gamma$ is a solution of the equations of motion, $\gamma$ is a field line of $\vec{S}$: $\gamma$ is tangent to $\vec{S}$ at all points. We can write $d\vec{\gamma} = \vec{S} dt$. The triple product, in this case, is zero regardless of what $d\vec{\lambda}$ is. We can reason the other way: we look for a line $\gamma$ such that the triple product is zero regardless of what $d\vec{\lambda}$ is. The only way for that to happen is if $\gamma$ is tangent to $\vec{S}$ at each point, which means $\gamma$ is a solution of the equations of motion. This means that the paths that make the line integral of the vector potential stationary are those and only those that solve the equations of motion. This recovers the principle of stationary action in Hamiltonian form
\begin{equation}
\tag*{(\ref{sdof_stationary_action})}
	0 =\delta \int_{\gamma} \vec{\theta} \cdot d\vec{\gamma} = - \iint_{\Sigma} \vec{S} \cdot d\vec{\Sigma} 
\end{equation}

We want to stress that this variational principle applies to any divergence-free field. For example, for a magnetic field we could write:
\begin{equation}
	\iint_{\Sigma} \vec{B} \cdot d\vec{\Sigma} = \delta \int_{\gamma} \vec{A} \cdot d\vec{\gamma}.
\end{equation}
We would find that the field lines of $\vec{B}$ are the lines and only the lines for which the integral of the vector potential is stationary. Similarly, we would find that the streamlines of an incompressible fluids are the lines and only the lines for which the integral of the vector potential of the flow is stationary. While these properties may be of limited practical use, the geometric insight is the same.

For (vi), if the assumption \ref{assum_kineq} of kinematic equivalence  holds, we can from a trajectory in space $q(t)$ reconstruct the state at any time. That is, if we are given $q(t)$ we can reconstruct the full evolution $\gamma(t)$ in the extended phase space.

We saw before that the evolution must be differentiable, which means $q(t)$ must also be differentiable. Therefore the velocity $\dot{q} = \frac{dq}{dt}$ is defined at every instant. Using assumption \ref{assum_detrev}, we know that each evolution is fully identified by position and momentum at a given time. Therefore the space of all possible evolutions is two-dimensional. Given that for each evolution $\gamma(t)$ there is one and only one spatial trajectory $q(t)$, the space of all spatial trajectories must also be two-dimensional and therefore fully identified by position and velocity at a given instant. Therefore there must be a map $p(q, \dot{q}, t)$ that allows us to reconstruct the momentum at a given time by knowing position and velocity at that time.

We can use this map to express the Hamiltonian form of the principle of stationary action only in terms of position and velocity. We have
\begin{equation}
\tag*{(\ref{sdof_Lagrangian})}
	\begin{aligned}
		\delta \int_\gamma \vec{\theta} \cdot d\vec{\gamma} &= \delta \int^{t_2}_{t_1} \vec{\theta} \cdot \frac{d\vec{\gamma}}{dt} dt \\
		&= \delta \int^{t_2}_{t_1} \left(\theta_q \frac{dq}{dt} + \theta_p \frac{dp}{dt} + \theta_t \frac{dt}{dt}\right) dt \\
		&= \delta\int^{t_2}_{t_1} \left(p \frac{dq}{dt} + 0 \frac{dp}{dt} - H \frac{dt}{dt}\right) dt \\
		&= \delta \int^{t_2}_{t_1} \left(p \frac{dq}{dt} - H\right) dt \\
		&= \delta \int^{t_2}_{t_1} L(q, \dot{q}, t) \, dt = 0
	\end{aligned}
\end{equation}
We recognize the principle of stationary action in standard form.

Note that the argument can proceed in the opposite way. If we have a system for which the principle of stationary action yields a unique solution, then we can write $p=\frac{\partial L}{\partial \dot{q}} = p(q, \dot{q}, t)$ which means we are able to reconstruct the momentum, and therefore the state, at each time. The system satisfies \ref{assum_kineq}. Moreover, we can find a unique Hamiltonian such that the solutions of the equations of motion are exactly those paths that satisfy the principle of stationary action. Therefore the system satisfies \ref{assum_detrev} as well.

Before proceeding to the generalization, a few observations are in order.
\begin{itemize}
	\item We stress that the vector potential $\vec{\theta}$ is not uniquely defined, and therefore neither are the action nor the Lagrangian. Giving precise physical characterizations of those two objects, then, is in our view a hopeless task.
	\item We found that the Hamiltonian $H$ is the time component of the vector potential $\vec{\theta}$, not a scalar. This relativistic aspect emerges naturally out of the assumptions.
	\item One can, in principle, write Lagrangian systems for which the action principle does not yield a single solution (e.g. for $L=0$ all paths are stationary). Consistently, assumption \ref{assum_detrev} fails.
\end{itemize}

\subsection*{Introduction to differential topology}

>>>>

\documentclass[fleqn,10pt]{wlscirep}
%\documentclass[10pt,twocolumn, nofootinbib]{revtex4-2}
%\documentclass[aps,pra,10pt,twocolumn,floatfix,nofootinbib]{revtex4-1}
%\documentclass[10pt,twocolumn,letterpaper]{article}

\setlength\columnsep{22pt} %To try to make line numbers work in 2-column format

\usepackage{assumptionsofphysics}
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{
	colorlinks=true,
	citecolor=blue,
	urlcolor=blue,
	linkcolor=blue
}
\urlstyle{same}
\usepackage{lineno}
\frenchspacing

@@

\begin{document}
\linenumbers

\title{Geometric and physical interpretation of the action principle}
\author[1]{Gabriele Carcassi}
\author[1,*]{Christine A. Aidala}
\affil[1]{Physics Department, University of Michigan, Ann Arbor, Michigan 48109 USA}

\affil[*]{caidala@umich.edu}

%\author{Gabriele Carcassi, Christine A. Aidala}
%\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}

@@

\section*{Introduction}

While the principle of stationary action is regarded by many as one of the most important tools in physics, its physical meaning is not completely clear.\cite{hamilton1834general, heaviside1903, feynman1942principle, brenier1989least, rojo2018principle} First of all, the typical characterization of the Lagrangian as the difference between kinetic and potential energy fails even for simple systems, like a charged particle under a magnetic field. Moreover, the Lagrangian for a system is not uniquely defined, making the actual value of the action for a path not directly physically significant.  We are left to wonder: what exactly is the action and why is it stationary for actual trajectories?

@@

The mathematics needed to run the argument is well established\cite{souriau1970structure, abraham1978foundations,arnold1989mathematical, marsden1999introduction}. The two aspects that have been sorely lacking, and that we provide, are (1) a clear geometric interpretation of the action principle and (2) a tight connection between the math and the physics it represents. [This has been the hardest problem to solve, as the modern mathematical concepts and notation have departed from physical intuition. We will therefore use, instead of the standard notation of differential geometry, an extended version of the one physicists use in general relativity.] To make the result accessible to the widest audience, we first cover the case of a single degree of freedom using standard vector calculus. We then proceed to the case of multiple independent degrees of freedom using tools from differential geometry. The main article will presents all the key points needed to follow the argument and its physical and geometric meaning, leaving the mathematical details and calculations to the Supplementary Information.

\section*{One degree of freedom}

@@

Under the assumption that
\begin{align}\label{assum_detrev}
	\tag*{(DR)}
	\parbox{2.8in}{\emph{the system undergoes deterministic and reversible evolution}}
\end{align}
we can define a displacement vector field
\begin{equation}\label{sdof_displacement}
\begin{aligned}
	\vec{S} &= \left[ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right] \\
	&= S^a e_a = \frac{d\xi^a}{dt} e_a .
\end{aligned}
\end{equation}
that describes how states move in time. [Where possible, we will be writing the same expression in both vector calculus and component notations.] In dynamical system literature, this is referred to as the vector field of the dynamical system. The time component of the displacement vector field is constrained, as we have
\begin{equation}\label{sdof_time_constraint}
	S^t=\frac{dt}{dt}=1.
\end{equation}

If assumption \ref{assum_detrev} is valid, we expect the flow of states through a closed surface to be zero: as many states flow in as flow out of the region. Alternatively, if we assign a probability, or probability density, to each trajectory, the assumptions requires that probability not to change, so integrating the probability over a closed surface must yield zero. However we see it, assumption \ref{assum_detrev} means the field is divergence-free.  [Given that this is a three dimensional space, we can use the standard tools of vector calculus.] That is, 
\begin{equation}\label{sdof_div_free}
	\nabla \cdot \vec{S} = \partial_a S^a = 0.
\end{equation}

@@

Because the displacement field must satisfy \ref{sdof_time_constraint}, without loss of generality we can set
\begin{equation}\label{sdof_potential_expression}
\begin{aligned}
	\vec{\theta} &= [p, 0, -H(q,p,t)] \\
	&= p e^q - H(q,p,t) e^t,
\end{aligned}
\end{equation}
where $H$ is a suitable function of $q$, $p$ and $t$. The potential $\vec{\theta}$ is closely related to the canonical one-form of symplectic geometry and the contact form of contact geometry. By applying definition \ref{sdof_displacement} and expanding \ref{sdof_displacement_potential} with \ref{sdof_potential_expression}, we have
\begin{equation}\label{sdof_Ham_eq}
	\left[ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right] = - \nabla \times \vec{\theta} = \left[ \frac{\partial H}{\partial p},-\frac{\partial H}{\partial q}, 1 \right],
\end{equation}
which yields Hamilton's equations. Note that the argument works in reverse: any Hamiltonian system with one degree of freedom yields a divergence-free displacement field, and therefore satisfies \ref{assum_detrev}.

@@

We now turn to constructing the principle of stationary action. As illustrated in Fig. \ref{fig_action}a, take a path $\gamma$ with endpoints $A$ and $B$, not necessarily a solution of the equations of motion. Then take a variation $\gamma'$ of that path and identify a surface $\Sigma$ between them. We can ask: what is the flow of the displacement field $\vec{S}$ through $\Sigma$? Because $\vec{S}$ is divergence-free, the flow through $\Sigma$ will depend only on the contour, therefore the question is well posed. Using Stokes' theorem, we find
\begin{equation}\label{sdof_action}
\begin{aligned}
	- \iint_{\Sigma} \vec{S} \cdot d\vec{\Sigma} &= \iint_{\Sigma} \left( \nabla \times \vec{\theta} \right) \cdot d\vec{\Sigma} \\
	&=  \oint_{\partial \Sigma = \gamma \cup \gamma'} \vec{\theta}  \cdot d\vec{\gamma}  \\
&= \int_{\gamma} \vec{\theta} \cdot d\vec{\gamma} - \int_{\gamma'} \vec{\theta} \cdot d\vec{\gamma}' \\
	&= \delta \int_{\gamma} \vec{\theta} \cdot d\vec{\gamma}.
\end{aligned}
\end{equation}

@@

\section*{Multiple degrees of freedom}

@@

The flow of $\vec{S}$ will still be divergence-free, but this property only tells us that the total number of states is conserved. We also need to assume that
\begin{align}\label{assum_indep}
	\tag*{(IND)}
	\parbox{2.8in}{\emph{the degrees of freedom are independent.}}
\end{align}
This means that, at each moment in time, the total number of states (i.e. the volume) of a parallelepiped will be the product of states identified by each degree of freedom (i.e. the areas). To be able to quantify the number of states identified by each degree of freedom, we introduce the rank-2 tensor (more precisely a two-form)
\begin{equation}\label{mdof_form}
	\omega = \omega_{ab} \, e^a \otimes e^b,
\end{equation}
which we call the state counting form. [Note that this approach is consistent with statistical mechanics\cite{peliti2011statistical}.]

@@

The form $\omega$ will need to be anti-symmetric
\begin{equation}\label{mdof_form_antisymm}
	\omega(\vec{v}, \vec{w}) = - \omega(\vec{w}, \vec{v})
\end{equation}
as the parallelogram identified by $\vec{v}$ and $\vec{w}$ in that order will be the same as the one identified by $\vec{w}$ and $\vec{v}$ with opposite orientation. The form $\omega$ will also be closed, meaning 
\begin{equation}\label{mdof_closed_form}
	\oiint_\Sigma \omega(d\Sigma) = 0
\end{equation}
over all closed contractible surfaces. [In our notation, $d\Sigma$ represents the infinitesimal surface element of integration, which is the argument of the form.] This stems from \ref{assum_indep}. Suppose you take a surface and translate it along another independent variable. The count of states cannot depend on the independent variable. Therefore if we construct any parallelepiped at equal time, two opposite sides will contain the same number of states with opposite orientation, and therefore the integral of $\omega$ over the whole surface will be zero. Those familiar with differential geometry will recognize $\omega$ as the symplectic form of symplectic manifolds and of the symplectic bundles of contact manifolds.

Since $\omega$ is closed, in every contractible region it can be expressed as the exterior derivative of a covector $\theta$:
\begin{equation}\label{mdof_form_potential}
	\begin{aligned}
		\theta &= \theta_a e^a = \theta_{q^i} e^{q^i} + \theta_{p_i} e^{p_i} + \theta_t e^t \\
		\omega &= - \partial \wedge \theta = - \left( \partial_a \theta_b - \partial_b \theta_a \right) e^a \otimes e^b
	\end{aligned}
\end{equation}
The minus sign is required to match conventions. [We write the exterior derivative as $\partial \wedge \omega$ instead of $d \omega$ so that $d$ is only used for infinitesimal line or surface elements.]

@@

Similarly to the single d.o.f., without loss of generality we can set
\begin{equation}\label{mdof_potential_expression}
\begin{aligned}
	\theta &= [p_i, 0, -H] \\
	&= p_i e^{q^i} - H e^t.
\end{aligned}
\end{equation}
and calculate the components of the counting form
\begin{equation}\label{mdof_form_components}
	\omega_{ab} = (-\partial\wedge\theta)_{ab} = \begin{bmatrix}
		0 & \delta^j_i & \partial_{q^i} H \\
		-\delta^i_j & 0 & \partial_{p_i} H \\
		-\partial_{q^j} H & -\partial_{p_j} H & 0
	\end{bmatrix}.
\end{equation}

@@

\section*{Discussion and conclusions}

@@

Another interesting element is that a version of the principle of stationary actions holds even if \ref{assum_kineq} does not apply. In this case the ``Lagrangian'' $L(q,\dot{q},p,t)$ depends on conjugate momentum as well. That is, the principle holds on all Hamiltonian systems, whether or not they admit a Lagrangian. Conversely, the cases where the Lagrangian does not admit a Hamiltonian are exactly the cases where the principle of stationary action fails to yield a unique solution, thus does not hold. These are the cases where the Lagrangian is not hyperregular. Therefore the principle of stationary action holds for all Lagrangian systems that admit a Hamiltonian, but also for all Hamiltonian systems whether of not they admit a Lagrangian. In this sense, the action principle is better understood as a feature of Hamiltonian mechanics, instead of Lagrangian mechanics\cite{souriau1970structure, arnold1989mathematical}.

@@

\vspace{-1mm}
\section*{Data availability}
All data generated or analysed during this study are included in this published article and its supplementary information files.

\vspace{-1mm}
\section*{Acknowledgments}
This paper forms part of the ongoing \textit{Assumptions of Physics} project \cite{aop-book}, which aims to identify a handful of physical principles from which
the basic laws can be rigorously derived. The authors would like to thank Alejandro Uribe-Ahumada and Daniel Burns for helpful discussions.

@@
